Towards a wide-coverage grammar for Swedish using GF

### 2011-12-29 doc/report/report.tex
report: refs, only pics missing

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{ amssymb }
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{parsetree}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{tipa}
\usepackage{textcomp}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

Further, thanks to my excellent 
%(~~devoted~~) 
supervisor Ramona Enache for all her help and guidence 
%(encouragement)
in every phase and all aspects of the work.
Thanks to Elisabet Engdahl for sharing her seemingly unlimited knowledge
of Swedish grammar. She has also has acted as a second supervisor, and given me
very helpful comments and suggestions. Thanks to Aarne Ranta for all his great
ideas and for letting me do this project.

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and represents a formalism
%equivalent to Parallel Multiple Context-Free Grammars (Seki \& al. 1991), an instance of
%generalized context free grammar (GCFG) it is hence
stronger than mildly context-free grammars. The framework has  hence the same expressiveness
as Tree Adjoining Grammars (Joshi,1975) and Head Grammars (Pollard,1984).

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pgf-format by the program evince.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\label{fig:shell}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.22\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree.
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{``the"},
and the determiner is therefore shown in the parse tree.\\

%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language, a limited language used in
a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems. The formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF can be used for describe dialog grammars see \cite{talk}, and
the framework is further used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

\subsection{Writing a GF grammar}
\label{sec:writegf}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
An abstract grammar can be implemented by a set of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as f-structures and c-structures in Lexical Functional
Grammar (Bresnan and Kaplan, 1982).

@@

In the TAG formalism (Joshi,1975), there are projects on getting open source, wide-coverage grammars
for English and Korean, but, to our knowledge, not for Swedish.  \\

The ParGram project aims at making wide coverage grammars using the Lexicon Functional Grammar approach
(Bresnan and Kaplan, 1982).
The grammars are implemented in parallel in order to coordinate the analyses of
different languages and there are now grammars for English, German, Japanese and Norwegian.

@@

\section{Implementation}
%>> syntactic information too. Mention that GF does not have this, only lexical or
%>> semantic. Not 1-1.
%TODO when??
The basic algorithm for importing Saldo was implemented by Angelov (2008)
and  produces code for a GF lexicon.
For each word in Saldo, it decides which forms should be used as input
to the the GF smart paradigms. For a verb, this will in most cases mean giving
the present tense form, see figure \ref{fig:saldoknyt}. \\

@@

\verb-mkV "knyter" "knöt" "knutit"-\\

@@

%compared to the total size of Saldo this is .
When testing the coverage of Talbanken,
we found that there are around 2500 word forms still missing, excluding the ones
tagged as names and numbers. This number may seem very high, but %considering that
4/5 of the word forms are compounds and when performing the inteded parsing,
these an additional analysis before being looked-up in the lexicon. We
should also take into consideration that 
we cannot automatically find out how many actually stem from the same word, or
how many abbreviations that are present (Talbanken also contains a small number of 
spelling errors, which probably are enumerated among our missing words). The majority
of the missing words are only used once.\\

@@

The function \verb-SlashV2a- lets you create a \verb-VPSlash-, a verb phrase missing a
noun phrase \cite{gazdar}. When we combine the \verb-SlashVP- with the object we
get a complete verb phrase.
Both the verb and its complements are stored in the verb phrase category, \verb-VP-.
%A simple idea would let the \verb-VP- consist of one field where the verb and object are
%glued together. But we also want to be able to express sentencens like
%``Du ser inte mig" and ``Mig ser du inte", where the object is separated from the verb.
The \verb-VP- category resembles the Didrichsen's field model.
%it has multiple fields that may be put in the different orders.
There are fields
for negation, different types of adverbs, objects, the finite verb and an infinite verb.
The fields may in turn be tables showing the different forms of the component.\\

@@

In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med")) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (``den" and ``honom").
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{`You look at the boy.'}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=90mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=90mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}

@@

Like verb phrases, a clause may also be missing an object and then has the type
\verb-ClSlash-. The \verb-ClSlash- is formed by a \verb-VPSlash- which is 
given a subject.
This is a convenient way to form questions, relative clauses
and topicalized clauses, as in figure \ref{fig:clause2}.\\

@@

%TODO
%>> change dot file to have VP NP in correct order

@@

There are also restrictions on the real subject, which is not allowed to be in
definite form.
\enumsentence{\begin{tabular}{llllll}
*Det     & sitter     & \textbf{den}& \textbf{fågeln}& på       & taket. \\
%& *Det sitter \textbf{fågeln} på taket.\\
\emph{it}& \emph{sits}& \emph{the}  & \emph{bird}    & \emph{on}& \emph{the roof}\\
%& ``There are the bird sitting on the roof'' \\
\end{tabular}}
Neither is sentence (\ref{sent:dumfagel}) correct.
\enumsentence{\begin{tabular}{llllll}
*Det& sitter& \textbf{min}& \textbf{fågel}& på& taket\\
\emph{it}&\emph{sits}&\emph{my}&\emph{bird}&\emph{on}&\emph{the roof}\\
%``There are my bird sitting %on the roof''\\
\end{tabular}
}\label{sent:dumfagel}
In order to implement this, we utilized the 
%The distinction lays in the determiner. The
different types of determiners shown in figure \ref{tab:detdef}. 
For postverbal subjects, the determiner must be of type \verb-DIndef-,
that requires both the noun and the adjective to be indefinite.
%The quantifier must have determiner must havenoun must be in indefinite form,
%even though some determiners are allowed.
%\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
%a. & Det sitter några fåglar på taket    & & b. & Det sitter fåglarna på taket\\
%   & ``There are some birds on the roof''& &    & ``There are the birds on the roof''\\
%\end{tabular}}\label{sent:faglar}
%Sentence \ref{sent:gamlafagel}b exemplifies a determiner that uses the noun in indefinite form,
%but the adjective in definite form.
%\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
%a. & Det sitter en gammal fågel på taket.& & b. & * Det sitter min gamla fågel på taket.\\
%   & ``There are some birds on the roof''& &    &~ ``There is my old bird on the roof.''\\
%\end{tabular}}\label{sent:gamlafagel}
%
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
To form a clause with formal subject, we hence need to inspect the determiner.
\begin{verbatim}
FormalSub : SimpleVP -> Det -> CN -> Cl ;
\end{verbatim}
The function combines a verb phrase, a determiner and a noun phrase to a clause.
If the determiner does not fulfill the requirements stated above, the clause is
put to NONEXIST.
This works well for parsing, but leads to problems if the grammar is used
for random generation. The solution is thus not ideal,
but since the definiteness of noun phrases or determiners cannot be seen on the
type level, it is not known until runtime whether the determiner is accepted
in the subject.

@@

\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
%\enumsentence{Han ger sina pengar till sina syskon.\\
%He gives {\sc self's} money to {\sc self's} children.}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
a. & Han är längre än sin kompis. & & b. &Han är här oftare än sin kompis.\\
   & He is taller than {\sc self's} friend.
   & & & He is here more often than {\sc self's} friend.
\end{tabular}}\label{ex:ReflAP} \label{ex:ReflAdv}
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Hon tyckte om skolan och alla sina elever. & & b. & Han såg sina få böcker och sin penna.\\
   & She liked the school and all {\sc self's} students. &&&
 He saw {\sc self's} few books and {\sc self's} pencil.\\ 
\end{tabular}}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in %, it is hence not
%allowed as subject. The antecedent must not be out of scope for the closest finite verb,
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antencedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar should not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar på tåget.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Hon och sin kompis läser en bok.\\
She and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser på sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

@@

The focusing adverbs are usually not combined with modal verbs, the copula or temporal auxiliries. 
% since the modal verbs rarely are focused.
\eenumsentence{\item[a]
\shortexnt{5}
{?&Han & bara&  är&  dum}
{&\emph{he} &\emph{ only} & \emph{ is}& \emph{ stupid}}
\item[b]
\shortexnt{5}
{?&Han & bara&  har&  sovit}
{&\emph{he} &\emph{ only} & \emph{ has}& \emph{ slept}}}
We have however chosen to allow this, since it is grammatically correct although semantically
dubious, and since \emph{``Han har bara sovit"} is covered by the rules using focusing
adverbs as \verb-AdV-.
%The semantic aspect of focusing adverbs remains rather unclear in this
%representation. 
%Discussion: should han åt bara ett äpple be (bara ett äpple) focusing an NP?
%Then what about: han har bara sovit, han har bara ätit ett äpple.\\

@@

Some more information about VP are given by the tag \verb|VG| (\emph{verb group}),
which groups an inifinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat version, and the results
get slightly better when using the flat version. \\

@@

Our final goal is to parse Talbanken, but considering that no syntactic theory is yet capable
of covering a whole natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish.
Following methods for making the parser robust as described in section \ref{sec:future}
we still believe that it is possible to achieve the goal. 
\\

>>>>

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{ amssymb }
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{parsetree}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{tipa}
\usepackage{textcomp}
%\usepackage{apacite}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

Further, thanks to my excellent 
%(~~devoted~~) 
supervisor Ramona Enache for all her help and guidance 
%(encouragement)
in every phase and all aspects of the work.
Thanks to Elisabet Engdahl for sharing her seemingly unlimited knowledge
of Swedish grammar. She has also has acted as a second supervisor, and given me
very helpful comments and suggestions. Thanks to Aarne Ranta for all his great
ideas and for letting me do this project.

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and represents a formalism
%equivalent to Parallel Multiple Context-Free Grammars (Seki \& al. 1991), an instance of
%generalized context free grammar (GCFG) it is hence
stronger than mildly context-free grammars. The framework has  hence the same expressiveness
as Tree Adjoining Grammars \cite{tag} and Head Grammars \cite{hg}.

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pgf-format by the program evince.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree.
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
For example, the definiteness and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{``the"},
and the determiner is therefore shown in the parse tree.\\
%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage
\begin{wrapfigure}{l}{0.35\textwidth}
\includegraphics[width=30mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}

The scope of GF grammars is usually controlled language, a limited language used in
a restricted domain. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities as well as ambiguities arising from multilinguality will
remain, but can be controlled more easily. %Example of domainspecific word?

The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems. The formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF can be used for describe dialog grammars see \cite{talk}, and
the framework is further used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

\vspace{10mm}
\subsection{Writing a GF grammar}
\label{sec:writegf}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
An abstract grammar can be implemented by a set of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as f-structures and c-structures in Lexical Functional
Grammar \cite{lfg}.

@@

In the TAG formalism \cite{tag}, there are projects on getting open source, wide-coverage grammars
for English and Korean, but, to our knowledge, not for Swedish.  \\

The ParGram \cite{pargram} project aims at making wide coverage grammars using
the Lexical Functional Grammar approach \cite{lfg}.
The grammars are implemented in parallel in order to coordinate the analyses of
different languages and there are now grammars for English, German, Japanese and Norwegian.

@@

\section{Implementation}
%>> syntactic information too. Mention that GF does not have this, only lexical or
%>> semantic. Not 1-1.
The basic algorithm for importing Saldo was implemented by Angelov (2008)
and  produces code for a GF lexicon.
For each word in Saldo, it decides which forms should be used as input
to the the GF smart paradigms. For a verb, this will in most cases mean giving
the present tense form, see figure \ref{fig:saldoknyt}. \\

@@

\begin{figure}[h]
\verb-mkV "knyter" "knöt" "knutit"-\\
\caption{Final output for the verb `knyta'}
\label{fig:saldoknyt3}
\end{figure}

@@

%compared to the total size of Saldo this is .
When testing the coverage of Talbanken,
we found that there are around 2500 word forms still missing, excluding the ones
tagged as names and numbers. This number may seem very high, but %considering that
4/5 of the word forms are compounds and when performing the intended parsing,
these an additional analysis before being looked-up in the lexicon. We
should also take into consideration that 
we cannot automatically find out how many actually stem from the same word, or
how many abbreviations that are present (Talbanken also contains a small number of 
spelling errors, which probably are enumerated among our missing words). The majority
of the missing words are only used once.\\

@@

The function \verb-SlashV2a- lets you create a \verb-VPSlash-. The name
\verb-SlashV2a- is inspired by categorical grammar and is a short
hand for \verb-VP \ NP- -- a verb phrase missing a
noun phrase..  When we combine the \verb-SlashVP- with the object we
get a complete verb phrase.
Both the verb and its complements are stored in the verb phrase category, \verb-VP-.
%A simple idea would let the \verb-VP- consist of one field where the verb and object are
%glued together. But we also want to be able to express sentencens like
%``Du ser inte mig" and ``Mig ser du inte", where the object is separated from the verb.
%TODO explain, diff from Gazdar
The \verb-VP- category resembles the Didrichsen's field model.
%it has multiple fields that may be put in the different orders.
There are fields
for negation, different types of adverbs, objects, the finite verb and an infinite verb.
The fields may in turn be tables showing the different forms of the component.\\

@@

In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med")) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (``den" and ``honom").
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{`You look at the boy.'}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=80mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=80mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}

@@

Like verb phrases, a clause may also be missing an object and then has the type
\verb-ClSlash-.
The \verb-ClSlash- is formed by a \verb-VPSlash- which is given a subject.
This is a convenient way to form questions, relative clauses
and topicalized clauses, as introduced by \cite{gazdar} (see figure \ref{fig:clause2}).\\

@@

@@

There are also restrictions on the real subject, which is not allowed to be in
definite form.
\enumsentence{\shortexnt{6}
{*Det     & sitter     & \textbf{den}& \textbf{fågeln}& på       & taket.}
%& *Det sitter \textbf{fågeln} på taket.\\
{\emph{it}& \emph{sits}& \emph{the}  & \emph{bird}    & \emph{on}& \emph{the roof}}
%& ``There are the bird sitting on the roof'' \\
}
Neither is sentence (\ref{sent:dumfagel}) correct.
\enumsentence{\shortexnt{6}
{*Det& sitter& \textbf{min}& \textbf{fågel}& på& taket}
{\emph{it}&\emph{sits}&\emph{my}&\emph{bird}&\emph{on}&\emph{the roof}}
%``There are my bird sitting %on the roof''\\
}\label{sent:dumfagel}
In order to implement this, we utilized the 
%The distinction lays in the determiner. The
different types of determiners shown in figure \ref{tab:detdef}. 
For postverbal subjects, the determiner must be of type \verb-DIndef-,
that requires both the noun and the adjective to be indefinite.
%The quantifier must have determiner must havenoun must be in indefinite form,
%even though some determiners are allowed.
%\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
%a. & Det sitter några fåglar på taket    & & b. & Det sitter fåglarna på taket\\
%   & ``There are some birds on the roof''& &    & ``There are the birds on the roof''\\
%\end{tabular}}\label{sent:faglar}
%Sentence \ref{sent:gamlafagel}b exemplifies a determiner that uses the noun in indefinite form,
%but the adjective in definite form.
%\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
%a. & Det sitter en gammal fågel på taket.& & b. & * Det sitter min gamla fågel på taket.\\
%   & ``There are some birds on the roof''& &    &~ ``There is my old bird on the roof.''\\
%\end{tabular}}\label{sent:gamlafagel}
%
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
To form a clause with formal subject, we hence need to inspect the determiner.
\begin{verbatim}
FormalSub : SimpleVP -> Det -> CN -> Cl ;
\end{verbatim}
The function combines a verb phrase, a determiner and a noun phrase to a clause.
If the determiner does not fulfill the requirements stated above, the clause is
put to NONEXIST.
This works well for parsing, but leads to problems if the grammar is used
for random generation. The solution is thus not ideal,
but since the definiteness of noun phrases or determiners cannot be seen on the
type level, it is not known until runtime whether the determiner is accepted
in the subject.

@@

\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
%\enumsentence{Han ger sina pengar till sina syskon.\\
%He gives {\sc self's} money to {\sc self's} children.}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
a. & Han är längre än sin kompis. & & b. &Han är här oftare än sin kompis.\\
   & He is taller than {\sc self's} friend.
   & & & He is here more often than {\sc self's} friend.
\end{tabular}}\label{ex:ReflAP} \label{ex:ReflAdv}
\eenumsentence{%\begin{tabular}[t]{@{}*{5}{l@{\ }}}
\item Hon tyckte om skolan och alla sina elever.\\
     She liked the school and all {\sc self's} students.
\item Han såg sina få böcker och sin penna.\\
 He saw {\sc self's} few books and {\sc self's} pencil.
}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in %, it is hence not
%allowed as subject. The antecedent must not be out of scope for the closest finite verb,
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar should not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar på tåget.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Hon och sin kompis läser en bok.\\
She and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser på sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

@@

The focusing adverbs are usually not combined with modal verbs, the copula or temporal auxiliaries. 
% since the modal verbs rarely are focused.
\eenumsentence{\item[a]
\shortexnt{5}
{?&Han & bara&  är&  dum}
{&\emph{he} &\emph{ only} & \emph{ is}& \emph{ stupid}}
\item[b]
\shortexnt{5}
{?&Han & bara&  har&  sovit}
{&\emph{he} &\emph{ only} & \emph{ has}& \emph{ slept}}}
We have however chosen to allow this, since it is grammatically correct although semantically
dubious, and since \emph{``Han har bara sovit"} is covered by the rules using focusing
adverbs as \verb-AdV-.
%The semantic aspect of focusing adverbs remains rather unclear in this
%representation. 
%Discussion: should han åt bara ett äpple be (bara ett äpple) focusing an NP?
%Then what about: han har bara sovit, han har bara ätit ett äpple.\\

@@

Some more information about VP are given by the tag \verb|VG| (\emph{verb group}),
which groups an infinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat version, and the results
get slightly better when using the flat version. \\

@@

Our final goal is to be able to parse unrestricted text, but considering that
no syntactic theory is yet capable
of covering a whole natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish.
We have therefor focused on constructions that are frequent in Talbanken, 
and following methods for making the parser robust as described in section \ref{sec:future}
we still believe that it is possible to achieve the goal. 
\\
